Exploring Compiler Optimization Techniques

### 2014-11-12 doc/paper.tex
Added introduction

%\hypersetup{}
%\lstset{numbers=left}
\title{Exploring SSA Optimization Techniques}
\author{A. D. Blom}
\date{}

\begin{document}
  \maketitle
  \begin{abstract}
In this era of global digitalization, the need for performant software is larger
than ever. For the production of fast software not only 
fast hardware and solidly written code are needed, but also a well-optimizing 
compiler. This paper explores some common optimization techniques based on an 
SSA intermediate language.
  \end{abstract}

  \section{Introduction}

  \section{Intermediate representation used}
The intermediate representation (IR) used here is an SSA implentation, 
partially based on the LLVM IR as described by~LLVM~\cite{llvm_ir}.
There are a few differences: the IR used here  uses a typesystem analogous to
C's, and uses a C-like syntax for textual 
representation as well.

@@

@@

>>>>

%\hypersetup{}
%\lstset{numbers=left}
\title{Exploring Compiler Optimization Techniques}
\author{A. D. Blom}
\date{}

\begin{document} \sloppy
  \maketitle
  \begin{abstract}
In this era of global digitalization, the need for performant software is larger
than ever. For the production of fast software not only 
fast hardware and solidly written code are needed, but also a well-optimizing 
compiler. This paper explores some common optimization techniques based on an 
SSA intermediate language.
  \end{abstract}

  \section{Introduction}
When a compiler translates code from a programming language to assembly languages,
it may apply a number of transformations to the source code, most often to make
programs faster (but also to enhance debugging capabilities for example). Doing
so is called code optimization.

Code optimization is most often done by first translating the source code
(as read by a \textit{lexer} and interpreted by a \textit{parser}) into an
\textit{intermediate form}, which is subsequently converted into assembly. Such
an intermediate form must be designed in such a way that it can be analyzed and
optimized easily.

The following C source code:

\begin{lstlisting}
int sub(int a, int b)
{
	return a - b;
}
\end{lstlisting}

May be converted (by the \textit{front-end}) into the following intermediate form
(this intermediate form is described in \ref{sec:irused}):

\begin{lstlisting}
global int (int, int)* @sub(int p0, int p1) {
L0:	int* t1 = alloca(int);
	int* t2 = alloca(int);
	store(t1, p0);
	store(t2, p1);
	int t3 = load(t1);
	int t4 = load(t2);
	int t5 = sub(t3, t4);
	ret(t5);
L1:	leave();
}
\end{lstlisting}

Which is subsequently optimized (by the \textit{middle-end}):

\begin{lstlisting}
global int (int, int)* @sub(int p0, int p1) {
L0:	int t1 = sub(p0, p1);
	ret(t1);
}
\end{lstlisting}

And converted into assembly (by the \textit{back-end}):

\begin{lstlisting}
	.globl	sub
sub:
	movl	%edi, %eax
	subl	%esi, %eax
	ret
\end{lstlisting}

It is then that the job of the compiler is done, and the assembler takes over,
and converts the assembly into numerical codes (\textit{"ones and zeroes"}):

\begin{lstlisting}
89 66 66 f8 f0 29 00 c3
\end{lstlisting}

One may ask oneself how the intermediate code is optimized from the first into
the significantly faster second format, and how this is implemented concretlely
in software. The \textit{acc} project was written to find out, and this article
describes some of the optimization techniques used by acc (and a few more).

  \section{Intermediate representation used}
  \label{sec:irused}
The intermediate representation (IR) used here is based on the \textit{static
single assignment} (SSA) principle: every temporary variable may be assigned to
only once. It is partially based on the LLVM IR as described by~LLVM~\cite{llvm_ir}.
There are a few differences: the IR used here  uses a typesystem analogous to
C's, and uses a C-like syntax for textual 
representation as well.

@@

\subsection{Common instructions}
A lot of the instructions are taken from LLVM. The non self-explanatory ones
behave as described below.
\subsubsection{alloca}
Allocates memory of the specified type, and returns a pointer to that memory.
The memory is deallocated when the function returns.
\subsubsection{split}
A conditional jump, \verb+split(a, b, c);+ jumps to \verb+b+ if \verb+a+, and
\verb+c+ if not.
\subsubsection{leave}
In a function returning \verb+T+, \verb+leave();+ is equivalent to
\verb+ret((T)undef);+. It returns from a function but returns no useful value.

@@

Instructions are represented as \verb+struct itm_instr+ pointers. They are
themselves a linked list: they contains pointers to the previous and next
instructions. They also link to their parent block, and have a list of their
operands. They contains a field of a strange type (\verb+itm_instr_id_t+) which
is an instruction identifier constant for each instruction type.

For instance, an instruction of the type \verb+add+ has a constructor called
\verb+itm_add()+. Its identifier can be obtained passing that function to the
\verb+ITM_ID()+ macro: \verb+ITM_ID(itm_add)+.

The structure looks roughly like this:

\begin{lstlisting}
struct itm_instr {
	struct itm_expr base;

	itm_instr_id_t id;
	struct itm_block *block;
	struct list *operands;
	struct itm_instr *prev, *next;
};
\end{lstlisting}
